# Binomial ~ Binomial

```{r include=FALSE}
library(tidyverse)
library(gtsummary)
library(foreign)
library(scales)
library(janitor)
library(flextable)
library(broom)
library(glue)

theme_set(
  jtools::theme_apa() +
    theme(
      plot.caption = element_text(hjust = 0)
    )
)
```

## Two Proportion Z-Test {#sec-tsztest}

The *Z*-test and chi-square test produce the same statistical significance result because they are algebraically identical.

The *Z*-test uses the difference in sample proportions, $\hat{d} = p_1 - p_2$, to test whether the difference in population proportions, $\delta = \pi_1 - \pi_2$, differs from an hypothesized difference, $d_0 = \pi_0 - \pi_1$. Alternatively, the *Z*-test is used to construct a $(1−\alpha)\%$ confidence interval around $\hat{d}$ to estimate $\delta$ within a margin of error, $\epsilon$.

The *Z*-test applies when the central limit theorem conditions hold so that the normal distribution approximates the binomial distribution.

* The sample is independently drawn, meaning random assignment (experiments) or random sampling without replacement from $n < 10\%$ of the population (observational studies),
* there are at least $n_i p_i >= 5$ successes and $n_i (1 - p_i) >= 5$ failures for each group $i$,
* the sample sizes are both $n_i >= 30$, and 
* the probability of success for each group is not extreme, $0.2 < \pi_i < 0.8$.

If these conditions hold, the sampling distribution of $\delta$ is normally distributed around $\hat{d}$ with standard error $se_\hat{d} = \sqrt{\frac{p_1(1 - p_1)}{n_1} + \frac{p_2(1 − p_2)}{n_2}}$. The measured values $\hat{d}$ and $se_\hat{d}$ approximate the population values $\delta$ and $se_\delta$. The *Z*-test statistic is $Z = \frac{\hat{d}}{se_{\hat{d}}}$.

Define a $(1 − \alpha)\%$ confidence interval as $\hat{d} \pm z_{\alpha / 2}se_\hat{d}$ or test the hypothesis of $d = d_0$ with test statistic $z = \frac{\hat{d} − d_0}{se_{d_0}}$ where $se_{d_0} = \sqrt{p^*(1 - p^*) \left(\frac{1}{n_1} + \frac{1}{n_2}\right)}$ and $p^*$ is the overall success probability.

## Fisher's Exact Test {#sec-fisherexact}

Fisher's exact test is an "exact test" in that the *p*-value is calculated exactly from the hypergeometric distribution rather than relying on the approximation that the test statistic distribution approaches $\chi^2$ as $n \rightarrow \infty$. The test is applicable in situations where

* the row totals $n_{i+}$ and the column totals $n_+j$ are fixed by study design (rarely applies), and
* the expected values of >20% of cells (at least 1 cell in a 2x2 table) have expected cell counts >5, and no expected cell count is <1.

The *p*-value from the test is computed as if the margins of the table are fixed. This leads under a null hypothesis of independence to a hypergeometric distribution of the numbers in the cells of the table ([Wikipedia](https://en.wikipedia.org/wiki/Fisher%27s_exact_test)). Fisher's exact test is useful for small *n*-size samples where the chi-squared distribution assumption of the chi-squared and G-test tests fails. Fisher's exact test is overly conservative (*p*-values too high) for large *n*-sizes.

The Hypergeometric density function is
$$f_X(k|N, K, n) = \frac{{{K}\choose{k}}{{N-K}\choose{n-k}}}{{N}\choose{n}}.$$

The density is the exact hypergeometric probability of observing this particular arrangement of the data, assuming the given marginal totals, on the null hypothesis that the conditional probabilities are equal.

## Case Study

This case study uses a data set from [Laerd](https://statistics.laerd.com/) and a second version modified to have small sample size. The first data set passes the *Z*-test requirements. The second (in parentheses), fails the sample sizes condition. 

```{r}
z_dat <- read.spss(
  "./input/test-of-two-proportions-individual-scores.sav",
  to.data.frame = TRUE
)

fisher_dat <- z_dat[seq(2, 100, 2),]
```

A researcher recruits `r nrow(z_dat)` (`r nrow(fisher_dat)`) patients who have a "high" classification of cholesterol and a poor lifestyle. The researcher randomly assigns `r z_dat %>% filter(intervention == "Drug") %>% nrow()` (`r fisher_dat %>% filter(intervention == "Drug") %>% nrow()`) of them to a drug intervention and `r z_dat %>% filter(intervention == "Lifestyle") %>% nrow()` (`r fisher_dat %>% filter(intervention == "Lifestyle") %>% nrow()`) to a lifestyle intervention. After six months, a doctor reclassifies the patients' cholesterol classification as either "high" or "normal".

```{r fig.height=3.5, fig.width=7.5}
#| code-fold: true
bind_rows(
  `Z-Test (n = 50, 50)` = z_dat,
  `Fisher (n = 25, 25)` = fisher_dat,
  .id = "set"
) %>%
  count(set, intervention, risk_level) %>%
  summarize(
    .by = c(set, intervention),
    p = sum(if_else(risk_level == "Normal", n, as.integer(0))) / sum(n),
    n = sum(n),
    se = sqrt(p * (1 - p) / n),
    ci_lwr = p - qt(.975, n - 1) * se,
    ci_upr = p + qt(.975, n - 1) * se
  ) %>%
  ggplot(aes(x = intervention)) +
  geom_col(aes(y = p), fill = "snow3", color = "snow4", width = 0.25) +
  geom_errorbar(aes(ymin = ci_lwr, ymax = ci_upr), color = "snow4", width = .125) +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(vars(fct_rev(set))) +
  labs(x = NULL, y = NULL,
       title = "Cholesterol Classification Improvement from High to Normal",
       subtitle = "Two data sets with different sample sizes.",
       caption = "Error bars show 95% CI.")
```

```{r}
#| code-fold: true
x <-
  z_dat |> 
  gtsummary::tbl_cross(
    row = intervention, 
    col = risk_level, 
    percent = "row", 
    label = list(intervention = "Intervention", risk_level = "Risk Level")
  )

y <- 
  fisher_dat |> 
  gtsummary::tbl_cross(
    row = intervention, 
    col = risk_level, 
    percent = "row", 
    label = list(intervention = "Intervention", risk_level = "Risk Level")
  )

gtsummary::tbl_merge(list(x, y), tab_spanner = c("Z-Test", "Fisher")) |>
  gtsummary::as_gt() |>
  gt::tab_header(title = "Summary statistics, Z-Test case, Fisher test case.") |>
  gt::tab_options(heading.align = "left")
```

### Conditions

The *Z*-test applies with the CLT conditions hold.

* the sample is independently drawn so the outcome of one observation does not influence another within the same group,
* both samples are randomly drawn from their respective populations so the results can be generalized,
* there are at least 5 successes (Normal) and failures (High) for each group $i$ so the normal approximation holds,
* each group has a fixed number of trials or observations,
* the population variances of the two groups are equal (typically true when sample sizes are large),
* the sample sizes for both groups are >=30, and 
* the probability of success for each group is not extreme, $0.2 < \pi_i < 0.8$.

The conditions hold for the *Z*-test data set, but not for the Fisher data set.

### Test

#### Z-Test {-}

There is no R function that performs a two-sample *Z*-test. Let's do it manually.

```{r collapse=TRUE}
dat_drug <- z_dat |> filter(intervention == "Drug")
dat_lifestyle <- z_dat |> filter(intervention == "Lifestyle")

n_drug <- nrow(dat_drug)
n_lifestyle <- nrow(dat_lifestyle)

p_drug <- mean(dat_drug$risk_level == "Normal")
p_lifestyle <- mean(dat_lifestyle$risk_level == "Normal")
p_pool <- mean(z_dat$risk_level == "Normal")

(se <- sqrt(p_pool * (1 - p_pool) * (1 / n_drug + 1 / n_lifestyle)))
(z <- (p_drug - p_lifestyle) / se)

(p_value <- pnorm(z, lower.tail = FALSE))
```

> `r nrow(z_dat)` patients with a high cholesterol classification were randomly assigned to either a drug or lifestyle intervention, `r n_drug` in each intervention. The test of two proportions used was the two sample *Z*-test. At the conclusion of the drug intervention, `r sum(dat_drug$risk_level == "Normal")` patients (`r percent(p_drug, 1)`) had improved their cholesterol classification from high to normal compared to `r sum(dat_lifestyle$risk_level == "Normal")` patients (`r percent(p_lifestyle, 1)`) in the lifestyle intervention, a difference in proportions of `r comma(p_drug - p_lifestyle, .01)`, *p* = `r comma(p_value, .0001)`.

The chi-square test of homogeneity (@sec-chisqhomogeneity) can also be used here.

```{r}
(chisq_test <- z_dat %>%
  tabyl(intervention, risk_level) %>%
  chisq.test(correct = FALSE))
```

> `r nrow(z_dat)` patients with a high cholesterol classification were randomly assigned to either a drug or lifestyle intervention, `r n_drug` in each intervention. The test of two proportions used was the chi-square test of homogeneity. At the conclusion of the drug intervention, `r sum(dat_drug$risk_level == "Normal")` patients (`r percent(p_drug, 1)`) had improved their cholesterol classification from high to normal compared to `r sum(dat_lifestyle$risk_level == "Normal")` patients (`r percent(p_lifestyle, 1)`) in the lifestyle intervention, a difference in proportions of `r comma(p_drug - p_lifestyle, .01)`, *p* = `r comma(chisq_test$p.value, .0001)`.

#### Fisher's Exact Test {-}

Use the Fisher exact test from the *janitor* package.

```{r}
(fisher_test <- fisher_dat %>%
  tabyl(intervention, risk_level) %>%
  fisher.test())
```

```{r include=FALSE}
fisher_drug <- fisher_dat |> filter(intervention == "Drug")
fisher_lifestyle <- fisher_dat |> filter(intervention == "Lifestyle")

p_fisher_drug <- mean(fisher_drug$risk_level == "Normal")
p_fisher_lifestyle <- mean(fisher_lifestyle$risk_level == "Normal")
```

> `r nrow(fisher_dat)` patients with a high cholesterol classification were randomly assigned to either a drug or lifestyle intervention, `r nrow(fisher_drug)` in each intervention. At the conclusion of the drug intervention, `r sum(fisher_drug == "Normal")` patients (`r percent(mean(fisher_drug$risk_level == "Normal"), 1)`) had improved their cholesterol classification from high to normal compared to `r sum(fisher_lifestyle$risk_level == "Normal")` patients (`r percent(mean(fisher_lifestyle$risk_level == "Normal"), 1)`) in the lifestyle intervention. Due to small sample sizes, Fisher's exact test was run. There was a non-statistically significant difference in proportions of `r comma(p_fisher_drug - p_fisher_lifestyle, .01)`, *p* = `r number(fisher_test$p.value, .0001)`.
